%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%A  xgaplatt.tex                XGAP documentation               Frank Celler
%%
%A  @(#)$Id: xgaplatt.tex,v 1.1.1.1 1996/12/11 12:39:53 werner Exp $
%%
%Y  Copyright 1995-1997,  Lehrstuhl D fuer Mathematik,  RWTH Aachen,  Germany
%%
%H  $Log: xgaplatt.tex,v $
%H  Revision 1.1.1.1  1996/12/11 12:39:53  werner
%H  Preparing 3.4.4 for release
%H
%H  Revision 1.3  1995/08/16  12:49:48  fceller
%H  fixed some more typos
%H
%H  Revision 1.2  1995/08/10  18:04:54  fceller
%H  fixed some typos
%H
%H  Revision 1.1  1995/08/09  10:49:05  fceller
%H  Initial revision
%%
\def\XGAP{{\sf XGAP}}
\Chapter{XGAP - Graphic Lattices}

{\XGAP} provides a graphical interface to the  lattice or partial lattice
of subgroups of groups.  For finitely  presented groups it gives you easy
access to the   low  index,  prime quotient   and   Reidemeister-Schreier
algorithms in order to build a partial  lattice interactively.  For other
types of groups  it provides easy access to  many of  the group functions
(for example, the normalizer, normal subgroups, and Sylow subgroups).

This chapter explains how to  use this interface.  Using various examples
the first  sections "The Subgroup  Lattice of $D_8$", "A Partial Subgroup
Lattice of $S_6$", "A Partial Subgroup Lattice of  $C_2$", and "A Partial
Subgroup  Lattice  of $K_3$" explain  most  features.  The later sections
give details about the various options and menus available.  This chapter
will  not describe  how  to write  your   own programs  using the graphic
extension supplied  by {\XGAP}, see  chapter "XGAP  - Graphic Extensions"
for details.

It is assumed that you have already started {\XGAP}.  On most systems you
do this by typing

|    user@host:~> xgap |

on  the  command line.   Ask your  system administrator  if this does not
work.  This   command  will create   a new window,   the so-called {\GAP}
window, in which {\GAP} is awaiting your input.   Depending on the window
system and window manager  you use, placing a new  window on your  screen
might be  done  automatically or might  require  you to use the  mouse to
choose a position for the  window and pressing the  left mouse button  to
place the window.

The small arrow   or cross you see on   your screen is  called a pointer.
Although the device used to move this pointer can be anything, a mouse, a
trackball, a glide-pad, or even something as exotic as a rat, we will use
the term mouse to refer to this pointer device.

In  case that some computation  takes  longer than expected, for instance
the low index and the prime quotient can be quite time consuming, you can
always interrupt  a computation  by  making the {\GAP}  window active and
pressing <CTR-C>.  Again, making  a  window active  is system and  window
manager dependent.   In most cases  you either have  to  move the pointer
inside  the {\GAP} window or you  have to click on the   title bar of the
{\GAP} window.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{The Subgroup Lattice of $D_8$}

This  section   gives  you an   example    on how  to    use the function
'GraphicLattice' (see "GraphicLattice" for  details), which will  display
the Hasse diagram of the subgroup lattice of a given group.

Using the dihedral group of  size $8$ as  example the following will show
you most features of the 'GraphicLattice' program.  This exercise is best
carried out in front of {\XGAP}, trying the various commands yourself.

First  you   have to define a  group   in {\GAP},  this example  uses the
dihedral group defined as permutation group.

|    gap> d8 := DihedralGroup(8);
    Group( (1,2,3,4), (2,4) )
    gap> d8.name := "d8";; |

'ConjugacyClassesSubgroups' computes and returns the conjugacy classes of
subgroups, so summing  up the sizes  of the classes   tells you how  many
elements the lattice has.

|    gap> Sum( List( ConjugacyClassesSubgroups(d8), Size ) );
    10 |

$10$ is small enough to use  'GraphicLattice' without painting the screen
black.  After you have entered

|    gap> s := GraphicLattice(d8);
    <graphic lattice> |

{\XGAP} will  open a window containing  a  new graphic  sheet, a menu bar
(menus are described below) above the graphic sheet and a title.  On most
systems the  title will be  either below the  graphic sheet or  above the
menu bar.  The dimension of the graphic sheet is fixed, changing the size
of the window will   *not* change the size   of  the graphic sheet,   see
"GraphicLattice,  Resize Menu"  how to resize  the  graphic sheet.  It is
possible  that the graphic  sheet is larger (depending  on the lattice it
might be much larger) than  the  window.  In  this  case the window  will
contain so-called scrollbars which allow you to select the portion of the
graphic sheet which will be displayed.

'GraphicLattice' then draws the  Hasse diagram onto the graphic sheet and
finally returns a {\GAP} record describing this sheet.

The   following initial  remarks  can    be  made  about  the   graphical
representation of the subgroup lattice\:

-- The vertex representing the trivial subgroup is labeled $1$.

-- Vertices representing subgroups of the same size are drawn at the same
height.  In  our example the subgroups that  belong to  the vertices $2$,
$3$, $4$, $5$ and $6$ all have size 2, and the subgroups of $7$, $8$, and
$9$ have  size $4$.   The default  behaviour is  to place a  vertex above
another one if the  size of the subgroup  represented by the first vertex
is larger     than  the size    of subgroup   of the  second,     but see
"GraphicLattice" for details.

-- Vertices  belonging to the same  conjugacy   class  are placed closely
together.  In our example the subgroups of $5$ and $6$ form one conjugacy
class.

The initial placement  of the vertices  chosen by 'GraphicLattice'  might
not be optimal or  you might want  to choose a  different one in order to
exploit certain  features  of the diagram.  It   is therefore possible to
move the vertices around using the mouse.

The mouse  together with the  left mouse button can be  used to  move and
select vertices. A selected  vertex is represented  by a thicker  circle,
colored red  if your screen  supports color.   For  example, in  order to
*move* vertex $6$ use  the mouse to  place the pointer inside  the circle
around $6$  and  press the *left*  mouse button.   Keep the  mouse button
pressed and start  moving the  mouse.   The vertex  will  now follow  the
pointer.  Because of the height restrictions given  by the size it is not
possible to move $6$ above $7$, $8$ or $9$ or  below $1$.  If you release
the  left mouse button vertex  $6$ will stay at  its current position and
the rest of  the conjugacy class (in  this example $5$)  will be moved to
this new position.

In order  to  *select* vertex $10$  place the  pointer inside  the circle
around  $10$, press the *left* mouse  button and  release it immediately.
Do not move the mouse while you hold down the left  mouse button.  Vertex
$10$ now has a slightly thicker  boundary and is red  if you have a color
screen.  There are  two different ways  to select more than  one  vertex,
see "A Partial Subgroup  Lattice of $S_6$" or "GraphicLattice,  Selecting
Vertices".

On the top of the window, above the graphic sheet,  you can see a list of
menu names\:\  <GAP>, <Resize>, <Subgroups>, and  <CleanUp>.  In order to
open any of these *pull  down menus* place the  pointer inside the button
containing   the menu name  and press  the  left  mouse button.  Keep the
button pressed.  A pull down menu will be shown and by moving the pointer
down you  can choose  a   menu entry.  By choosing    an entry and   then
releasing  the  mouse button  the entry   is  selected, the corresponding
function is executed and  the pull down menu is  closed.  If  you release
the mouse button while the pointer is outside the pull down menu the menu
is closed without selecting any entry.

Now select  <Relabel Vertex> from the  <CleanUp> menu.  If this  entry is
not available you have  failed to select   vertex $10$.  After  selecting
<Relabel Vertex> a small  dialog box is opened  asking for a label.  Type
in |D8| and press the <return> key or click on <OK>.  The label of vertex
$10$ will now be changed to $D8$.

In order to find  out which vertex  represents the centre of $D_8$, first
select vertex $D8$ and then the menu entry <Centres> from the <Subgroups>
menu.   In  case of  a  color screen, vertex $D8$   will  be selected and
colored  red, and vertex  $2$ will  be selected  and  colored green.  The
color green  indicates that vertex $2$  is  the result  of a computation.
There will also be a message in the {\GAP} window  saying that vertex $2$
represents the centre of the group belonging to vertex $D8$.

|    &I  Centre(D8) = 2 |

Most of  the menu entries  in <Subgroups> should be self-explanatory, for
details and  the   difference   between  <Closure>  and    <Closures> see
"GraphicLattice, Subgroups Menu".

If you have selected some  vertices (in the example  $D8$ and $2$ are now
selected), and you  want to investigate  the  subgroups corresponding  to
these vertices further in {\GAP},  the function 'Selected' will return  a
list of these subgroups.

|    gap> Selected(s);
    [ d8, Subgroup( d8, [ (1,3)(2,4) ] ) ] |

On the other hand, the functions supplied via the <Subgroups> menu are by
far not all functions applicable to groups.  In  order to show results of
a computation   in {\GAP} in  the diagram,   you  can use  'Select'.  The
function 'Select' allows you to  mark any series or  subgroup of $D_8$ in
the diagram.

For instance,  $D_8$ is  represented as  a permutation group,  so you can
compute a stabilizer chain in {\GAP}.

|    gap> b := Base(d8);
    [ 2, 1 ]
    gap> sc := [ d8, Stabilizer(d8,1), TrivialSubgroup(d8) ];
    [ d8, Subgroup( d8, [ (2,4) ] ), Subgroup( d8, [  ] ) ]
    gap> Select( s, last );
    [ <vertex "D8">, <vertex "3">, <vertex "1"> ] |

This stabilizer chain corresponds to the vertices $D8$, $3$ and $1$ which
will now be selected.  To summarize the  above\:\ the function 'Selected'
can  be  used to  transfer  information from the   diagram to {\GAP}, the
function 'Select' can be used to  transfer information from {\GAP} to the
diagram.

While pull down menus are bound to a menu button and  will be pulled down
as soon as  this button is  pressed, there is  another type of  menu, the
so-called *pop up menu*.  One of these pop up menus can be used to adjust
the size of the graphic sheet.  In order to adjust the left margin, place
the pointer  outside of any  vertex and press  the *right*  mouse button,
keep the button pressed.  A small menu will pop up.

|    Left Margin
    Right Margin
    Top Margin
    Bottom Margin |

Select <Left Margin>.   This will narrow  the graphic sheet,  setting the
new left margin  to the x coordinate of   your mouse click.   Repeat this
process with the right margin.  Depending on the window system and window
manager   narrowing  the graphic   sheet   might also   narrow the window
containing  the  graphic sheet.  If  not,  you  have to use  the standard
method of your window system/manager to change the size of this window.

In order to  finish this example, close  the  window by selecting  <close
graphic   sheet>  from  the  <GAP>  menu.    This will  close  the window
containing the Hasse diagram of $D_8$.

In this example you have learned, how to display the Hasse diagram of the
subgroup lattice of a group using  'GraphicLattice', how to use the mouse
to move and  select  vertices, how to select   a  menu entry  and  how to
transfer information between the Hasse diagram  and {\GAP} using 'Select'
and 'Selected'.

In  order to learn  more about  the menus  <GAP>, <Resize> and <CleanUp>,
which  were only mentioned  very briefly, see "GraphicLattice, GAP Menu",
"GraphicLattice, Resize Menu" and  "GraphicLattice, CleanUp Menu".   This
example  also didn\'t    discuss  two   additional  menus,  namely    the
<Information> and <Sheet  Menu>.  The  first is  discussed in "A  Partial
Subgroup Lattice  of $S_6$"  and "GraphicLattice, Information  Menu", the
latter in "GraphicLattice, Sheet Menu".

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{A Partial Subgroup Lattice of $S_6$}

This section investigates the subgroup lattice of $S_6$.

|    gap> s6 := SymmetricGroup(6);
    Group( (1,6), (2,6), (3,6), (4,6), (5,6) )
    gap> s6.name := "s6";;
    gap> cc := ConjugacyClassesSubgroups(s6);;
    gap> Sum( List( cc, Size ) );
    1455 |

As  there are $1455$ subgroups, displaying the whole lattice of subgroups
would not be helpful  because there are simply  too many.  Therefore this
example builds  only a partial subgroup  lattice.  We assume that you are
familiar with  the general  ideas,  mouse actions and menus,   which were
discussed in "The Subgroup Lattice of $D_8$".

In  order  to build  a  partial  lattice, use   'InteractiveLattice' (see
"InteractiveLattice") instead of  the  function 'GraphicLattice'.   After
you have entered

|    gap> s := InteractiveLattice(s6);
    <interactive graphic lattice> |

{\XGAP}  will open  a window  containing  a new  graphic  sheet with  two
connected vertices   labeled  $1$ and $2$.     Vertex  $1$ represents the
trivial subgroup and  vertex $2$ the group $S_6$.   Vertex $2$ is already
selected, so it will be red if your screen supports color.

In order  to  find all subgroups   of size $60$, we  cannot   not use the
<Subgroups> menu directly, so go back into the  {\GAP} window and extract
the conjugacy classes of <cc> whose representatives have size $60$.

|    gap> c60 := Filtered( cc, x -> Size(Representative(x)) = 60 );;
    gap> r60 := List( c60, Representative );
    [ Subgroup( s6, [ (1,3)(2,4), (1,6,4) ] ), 
      Subgroup( s6, [ (1,3)(2,4), (1,3,4)(2,5,6) ] ) ] |

Again,  use the  function 'Select'  to  add  these two subgroups  to your
partial lattice.  The Hasse diagram now contains four vertices.

|    gap> Select( s, r60 );
    [ <vertex "3">, <vertex "4"> ] |

The  new  vertices  are   selected automatically.    Selecting <Conjugate
Subgroups> from the <Subgroups> menu adds the complete conjugacy classes.
In order  to find out what  type of subgroups  we are  looking at,  use a
another kind of menu not discussed so far, namely the <Information> menu.
Place the pointer inside vertex  $3$, press the  *right* mouse button and
release it immediately.  This  will pop up a  new window, containing some
text describing  vertex $3$ (as  mentioned above, depending on the window
system   and window manager,  placing   this window on  the  screen might
require some interaction with the mouse).

|    Size            2^2*3*5
    Index           12
    IsAbelian       unknown
    IsCentral       unknown
    IsCyclic        unknown
    IsNilpotent     unknown
    IsNormal        unknown
    IsPerfect       unknown
    IsSimple        unknown
    IsSolvable      unknown
    Isomorphism     unknown |

Place the pointer on top of the entry  <Isomorphism> and press the *left*
mouse button.  After a while this entry is changed to

|    Isomorphism     A5 |

telling you that the subgroup represented by  vertex $3$ is isomorphic to
the alternating group  on   five symbols.  Select  <close>  to  close the
<Information>  menu.  Repeat this with vertex  $4$, you will see that the
subgroup of  vertex $4$ is also isomorphic   to $A_5$, however  these two
$A_5$ inside $S_6$ are  not conjugate in $S_6$.   The information menu is
described in detail in "GraphicLattice, Information Menu".

Now we want to compute the  normalizers of the  elements of the conjugacy
class containing the  subgroup of vertex  $3$.   You could either  select
vertex  $3$ and  then  <Normalizers>  and repeat   this process for   the
vertices $5$ to $9$, or you can first select the vertices $3$, $5$ to $9$
and  then select <Normalizers>.  But how  to select more than one vertex?
If you first select $3$ and  then $5$, vertex $3$  will get deselected as
soon as $5$ gets selected.  However, if you  select vertex $3$, place the
pointer inside vertex $5$, hold down the <SHIFT> key on your keyboard and
then select vertex $5$ using  the left mouse button,  vertex $5$ will  be
selected in addition to  vertex $3$.  Another  method to select more than
one  vertex is  to  use  the  rubber   band  to catch vertices  inside  a
rectangle.  Place  the pointer  left and  a  bit  higher than  vertex $3$
*outside*  any other vertex.   Press the *left*  mouse button and hold it
down.  Now, using  the mouse, move the  pointer right and slightly  below
vertex $9$.  You  see a rectangle, one corner  at your start position and
the other   following the pointer.   If  vertex $3$,  $5$ to $9$  are all
inside this rectangle, release the mouse  button.  Now these vertices are
selected.  Select <Normalizers> from the  <Subgroups> menu to compute and
display the normalizers.

Now select   vertex  $3$ and  $12$ and   compute   the intersection.  The
intersection is of size $10$.  Select this intersection and use  'Select'
to get a {\GAP} record describing the subgroup.

|    gap> l := Selected(s);
    [ Subgroup( s6, [ (2,3)(4,6), (1,2)(3,4) ] ) ]
    gap> u := l[1];
    Subgroup( s6, [ (2,3)(4,6), (1,2)(3,4) ] ) |

In order  to find out which subgroups  of the complete lattice  lie above
the subgroup <u> you can  use the following  sequence of {\GAP} commands.
<cc>,   which  was  defined above,   contains   all conjugacy  classes of
subgroups of $S_6$.

|    gap> l := List(cc, x->Filtered(Elements(x), y->IsSubgroup(y,u)));;
    gap> l := Concatenation(l);
    [ Subgroup( s6, [ (1,6,2,3,4), (1,4)(3,6) ] ), 
      Subgroup( s6, [ (1,6,2,3,4), (1,4)(3,6), (1,6,4,3) ] ), 
      Subgroup( s6, [ (1,3)(2,4), (1,6,4) ] ), 
      Subgroup( s6, [ (1,3)(2,6), (1,3,6)(2,4,5) ] ), 
      Subgroup( s6, [ (1,3)(2,4), (1,6,4), (4,6) ] ), 
      Subgroup( s6, [ (1,3)(2,6), (1,3,6)(2,4,5), (3,6,4,5) ] ), 
      Subgroup( s6, [ (1,5,2,4,6), (2,3)(5,6) ] ), s6 ]
    gap> Select( s, l ); |

There is another feature we have not seen yet.  Close the current graphic
sheet and start again with a fresh one.

|    gap> s := InteractiveLattice(s6); |

In order to compute a Sylow $2$ subgroup select <Sylow Subgroup> from the
<Subgroups> menu.   A small dialog box  will  pop up asking  for a prime,
type in  $2$ and press <return> or  click on <OK>.   Now select  this new
vertex $3$   representing the Sylow $2$  subgroup  and compute its normal
subgroups.  This is rather slow because the  function checks for each new
vertex if the corresponding  subgroup is conjugate to  an old one  of the
same size.  There  is however a  limit on the  index of the normalizer or
subgroup up to which this test is performed.

|    gap> s.limits.conjugates;
    100 |

If  'limit.conjugates' contains a   positive  number  the index  of   the
normalizer of the new subgroup is checked. If 'limit.conjugates' contains
a negative number, only the index of the  subgroup itself is checked.  If
the index  is less  than the  absolute  value then the conjugacy  test is
carried out.  Change this limit to $-10$.

|    gap> s.limits.conjugates := -10;
    -10 |

Select vertex $3$   and compute a  random conjugate  by selecting <Random
Conjugate> from the <Subgroups> menu.  Select this new vertex and compute
its  normal subgroups.  You will  see  that now  the vertices  of the new
subgroups appear  faster on  the   graphic sheet, however,  they  have  a
different shape.  They are  represented by a square  instead of a circle.
A square  is meant as  a danger sign,  because {\GAP} has not  checked if
subgroups associated  with square vertices  are  conjugate to any  of the
other subgroups.  Therefore these subgroups might be (and in this example
will be) conjugate to an old subgroup.

In order  to untangle the Hasse diagram you  can  move the new (selected)
vertices  en block.  Hold down  the <SHIFT> key and   move any of the new
vertices to the right.  All selected vertices will automatically be moved
as soon as you release the left mouse button.

This is  now the  end  of the investigation   of  the (partial)  subgroup
lattice   of  $S_6$, close the    graphic  sheet(s) using  <close graphic
lattice> of the <GAP> menu.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{A Partial Subgroup Lattice of $C_2$}

This  section  investigates the  following  finitely presented group, the
so-called Cavicchioli-group $C_2$.
$$
\langle a, b \;;\; aba^{-2}ba=b, (b^{-1}a^3b^{-1}a^{-3})^2=a^{-1}\rangle.
$$

The following {\GAP} commands define $C_2$.

|    gap> f := FreeGroup( "a", "b" );  a := f.1;;  b := f.2;;
    Group( a, b )
    gap> c2 := f / [ a*b*a^-2*b*a/b, (b^-1*a^3*b^-1*a^-3)^2*a ];
    Group( a, b )
    gap> c2.name := "c2";; |

We  again  assume that you are  familiar   with the general  ideas, mouse
actions  and menus,  which were  discussed  in "The  Subgroup Lattice  of
$D_8$" and "A Partial Subgroup Lattice of $S_6$".

In order  to build a  partial lattice of a  finitely presented group, you
again use the function  'InteractiveLattice'.  But if the first  argument
to 'InteractiveLattice' is a finitely presented group the available menus
are different  from the example in  the previous section.  After you have
entered

|    gap> s := InteractiveLattice(c2);
    <interactive graphic lattice> |

{\XGAP} will open a window  containing a new  graphic sheet.  Compared to
the  interactive  lattice of  a permutation   group  as described  in the
previous section, there are the following differences\:

-- There is only one  vertex instead of two.   This vertex labeled $1$ is
not the trivial  group, as in the last  two sections but  the whole group
$C_2$.  There is no vertex for the trivial subgroup (yet).

-- The shape of  vertex $1$ is not a  circle but a diamond.  Whenever you
see a diamond this  means that the subgroup  belonging to this vertex  is
normal in the whole group *and* {\GAP} knows this fact.

-- If you pull down the <Subgroups> menu, you will see  that this menu is
now   very different.   It gives  you   access to various algorithms  for
finitely presented  groups  but most of  the  entries  from the last  two
examples  are missing because most of  the {\GAP}  functions behind these
entries are not applicable to (infinite) finitely presented groups.

This  example will show  you how to prove that  $C_2$ is infinite.  First
look at the abelian invariants in order to see what the commutator factor
group is.   In  order  to  compute the abelian    invariants pop  up  the
<Information> menu.   This is done in exactly  the same manner  as in the
previous section.  Place the pointer inside vertex $1$, press the *right*
mouse button and release   it  immediately.  This <Information> menu   is
described  in   detail in  "InteractiveLattice  for FpGroups, Information
Menu".

|    Index                  1
    IsNormal               true
    Abelian Invariants     unknown
    Presentation           2 gens, 2 rels
    Coset Table            known
    Factor Fp Group        unknown |

This tells you what {\GAP} already knows  about the group associated with
vertex $1$.   In order to compute the  abelian invariants click onto this
line.  After a while this entry will change to

|    Abelian Invariants     perfect |

telling you  that $C_2$  is perfect.   So none   of the  <Subgroups> menu
entries <Abelian  Prime Quotient>,  <All Overgroups>,  <Conjugacy Class>,
<Cores>,  <Derived Subgroups>,  <Intersection>,   <Normalizer> or  <Prime
Quotient> will compute any new subgroups.

In order  to avoid accidents  the menu entries  <Abelian Prime Quotient>,
<All Overgroup>, <Conjugacy Class>, <Epimorphisms>, <Low Index Subgroups>
and  <Prime  Quotient> from the  <Subgroups>  menu are only selectable if
exactly one vertex is selected because the functions behind these entries
are in general quite time and space consuming.

Close the <Information> window and select <Low  Index Subgroups> from the
<Subgroups> menu.  A small dialog box will  pop up asking  for a limit on
the index.  Type in $12$ and press <return> or click on <OK>.  In general
it  is hard to say what   kind of index  limit  will still work, for some
groups even $5$ might be too  much while for others  $20$ works fine, see
also "LowIndexSubgroupsFpGroup".

{\GAP} computes $10$ subgroups  of index $11$  and $8$ subgroups of index
$12$.  If you now start to check the abelian invariants of the index $12$
subgroups you will  find out that all  subgroups represented by  vertices
$2$ to $9$  have  a finite  commutator  factor group except  the subgroup
belonging  to  vertex  $3$    which has  an infinite    abelian quotient.
Therefore the group $C_2$ itself is infinite.

Now we want to investigate $C_2$ a little further using {\GAP}.  Selected
vertices  $2$, $3$, and   $4$ and  switch   to  the {\GAP}  window.   Use
'Selected' to get the subgroups associated with these vertices.

|    gap> u := Selected( s );
    [ Subgroup( c2, [ a, b*a^2*b^-2, b*a*b^2*a^-1*b^-1*a^-1*b^-1,
                      b^4*a^-2*b^-2, b^2*a^3*b^-1*a^-1*b^-2 ] ), 
    Subgroup( c2, [ a, b^2*a*b^-1*a^-1*b^-1, b^3*a^-1*b^-1,
                    b*a*b*a^3*b^-1 ] ), 
    Subgroup( c2, [ a, b^2*a*b^-1*a^-1*b^-1, b*a^3*b^-2, b^4*a^-1*b^-3, 
                    b*a*b^3*a^-1*b^-1 ] ) ] |

'OperationCosetsFpGroup' computes for each of  these subgroups $u_i$  the
operation of  $C_2$ on its cosets. The  operation on $u_i$ is therefore a
permutation representation of the factor group $$C_2 / Core(u_i).$$ Using
'DisplayCompositionSeries' we can identify these factor groups.

|    gap> p := List( u, x -> OperationCosetsFpGroup( c2, x ) );;
    gap> for x  in p  do DisplayCompositionSeries(x);  Print("\n\n");  od;
    <G> (2 gens, size 95040)
     I M(12)
    <1> (0 gens)

    <G> (2 gens, size 660)
     I A(1,11) = L(2,11) ~ B(1,11) = O(3,11) ~ C(1,11) = S(2,11)
     I ~ 2A(1,11) = U(2,11)
    <1> (0 gens)

    <G> (2 gens, size 239500800)
     I A(12)
    <1> (0 gens) |

So $C_2$  contains the Mathieu group  $M_{12}$, the alternating  group on
$12$ symbols and $PSL(2,11)$ as  factor groups.  Therefore it would  have
been possible to   find vertex $3$ using  <Epimorphisms>  instead of <Low
Index Subgroups>.  Close  the graphic sheet by  selecting  the menu entry
<close graphic sheet> from the <GAP> menu and start with a fresh one.

|    s := InteractiveLattice(c2);
    <interactive graphic lattice> |

Select <Epimorphisms> from  the <Subgroups> menu.   This  pops up a  menu
similar to the <Information> menu  (see "InteractiveLattice for FpGroups,
Subgroups Menu").

|    Sym(n)
    Alt(n)
    Sym(n) > Alt(n)
    PSL(d,q)
    Library
    User Defined |

Select  <Library>, which  pops   up a file  selector.   Choose  the  file
|l2_11.grp| and click on <OK>.  After a while the entry will change to

|    Library (l2_11)      1 found |

telling  you,  that  {\GAP}  has   found $1$  epimorphism   (up to  inner
automorphisms  of  $PSL(2,11)$) from  $C_2$   onto $PSL(2,11)$.  Click on
<display> to create a  new vertex representing a  subgroup $u$ such  that
the   factor group of  $C_2  /  Core(u)$  is  isomorphic to  $PSL(2,11)$.
Instead of <Library> you could have chosen <PSL(d,q)>.

This  is now  the  end of  the investigation   of the (partial)  subgroup
lattice  of $C_2$, you have  seen  that $C_2$   is infinite and  contains
$M_{12}$, $Alt(12)$, and $PSL(2,11)$ as factor groups.  Close the graphic
sheet by selecting <close graphic sheet> from the <GAP> menu.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{A Partial Subgroup Lattice of $K_3$}

This  section  investigates the  following  finitely presented group, the
trefoil knot group $K_3$.
$$
    \langle a, b \;;\; aba = bab \rangle
$$

|    gap> f := FreeGroup( "a", "b" );
    Group( a, b )
    gap> k3 := f / [ f.1*f.2*f.1 / (f.2*f.1*f.2) ];
    Group( a, b )
    gap> s := InteractiveLattice(k3);
    <interactive graphic lattice> |

If you compute the   abelian invariants of  $K_3$  you will see  that the
commutator factor group  is isomorphic to the  infinite cyclic group.  If
you try  to compute the   derived subgroups  {\GAP}  will print  a  short
warning

|    &W  abelian quotient of 1 is infinite. |

because  'InteractiveLattice'  cannot   deal  with subgroups of  infinite
index.  So  use <Prime Quotient> to   compute a $2$-quotient  of $K_3$ of
class $4$ instead.  This creates a new vertices $2$, $3$, $4$ and $5$ for
the $2$-quotients of  class $1$ up to $4$  which however have a different
shape. They are diamonds inside a square.  A diamond again indicates that
the associated subgroup  is normal.  A square is  meant as a danger sign,
whenever you see a square, {\GAP} has not yet  computed a coset table for
this subgroup in the group.  Because no coset  table is known, {\GAP} has
*not  checked* if this  subgroup is  equal or conjugate   to any of these
other  subgroups nor has  it checked  any  inclusions.  So it is possible
that  there are two  vertices on  the sheet (where    at least one has  a
square) which   represent  the *same* subgroup.    We   will see  such  a
constellation now.

In order to further investigate $K_3$, compute the  subgroups of index at
most $4$.  There is one normal subgroup of  index $2$ belonging to vertex
$8$ and one of index $4$.  However, because the quotient  of $K_3$ by the
subgroup of vertex $2$ is  the largest elementary abelian $2$-quotient of
$K_3$, the subgroups belonging to vertex  $2$ and $8$ must  be equal.  In
order  to check this,  pop up the  <Information> menu for vertex $2$, and
compute a coset table.  As soon  as {\GAP} knows  a coset table it checks
this table against the other known coset tables. If it discovers that two
subgroups are equal, {\GAP} deletes   one vertex and updates the  partial
lattice.  In this example, {\GAP} deletes vertex $8$, and connects vertex
$9$ with $2$.

You can force this check for a set of vertices by selecting the vertices,
in our example  $3$, $4$ and $5$, and  then <Check Coset Tables> from the
<CleanUp> menu.  But keep in mind that  especially the results of a prime
quotient can have huge indices.

This is  now   the end  of the investigation   of  the (partial) subgroup
lattice of $K_3$,  close the graphic  sheet  by selecting <close  graphic
sheet> from the <GAP> menu.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{GraphicLattice}

'GraphicLattice( <g> )'

'GraphicLattice' creates a new graphic sheet containing the Hasse diagram
of the  subgroup lattice of <g>.  The  next next sections describe how to
select  and  move  vertices   and the  following  sections  describe  the
available menus.

|    gap> GraphicLattice( DihedralGroup(8) );
    <graphic lattice> |

'GraphicLattice( <g>, <width>, <height> )'

In   this form  'GraphicLattice'  creates  a   graphic sheet  of  initial
dimensions <width>    times <height>.  It is   however  still possible to
change these dimensions  later using either "GraphicLattice, Resize Menu"
or "GraphicLattice, Sheet Menu".

'GraphicLattice( <g>, \"prime\"\ )'

If you supply the parameter \"prime\", then a vertex will be placed above
another  vertex  if the   number  of prime factors   in  the size  of the
corresponding subgroup is larger than the number of  prime factors in the
size  of  the  subgroup belonging   to  the second  vertex.   The default
behaviour is to compare the sizes instead of number of prime factors.

|    gap> GraphicLattice( CyclicGroup(20) );;
    gap> GraphicLattice( CyclicGroup(20), "prime" );; |

'GraphicLattice( <g>, \"normal subgroups\"\ )'

If you supply the parameter  \"normal subgroups\", then only the  lattice
of normal subgroups of <g> will be drawn on the graphic sheet.

|    gap> GraphicLattice( DihedralGroup(8), "normal subgroups" );' |

It is possible to abbreviate \"prime\"\  and \"normal subgroups\"\ and to
supply one or both.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{GraphicLattice, Moving Vertices}

In order to move a vertex, place the  pointer inside the vertex using the
mouse, and press the *left* mouse button.  Hold the mouse button pressed,
and start moving  the pointer by moving  the mouse.  The  vertex will now
follow the pointer, but it is not possible to move the vertex higher than
a vertex  of bigger size  (or number of prime  factors, if \"prime\"\ was
given as parameter to 'GraphicLattice') or lower than a vertex of smaller
size (or number of prime factors).  As soon as you release the left mouse
button the vertex will stop following the pointer  and, if the vertex was
a member of a  conjugacy class, the remaining elements  of the class  are
moved.

If you hold  down  the <SHIFT> key before  moving  a vertex as  described
above then  not only is the  class  of the vertex   moved but all classes
containing at least one selected vertex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{GraphicLattice, Selecting Vertices}

Selected vertices are  represented by a slightly  thicker circle and,  if
your screen  supports color, are colored red   or green.  There  are five
different ways to select or deselect a vertex or a bunch of vertices.

Place the  pointer inside  a  vertex and press   the *left* mouse button.
Release   the button immediately   without  moving the  mouse.  This will
deselect all other vertices  and select this vertex.   If the vertex  was
already the only selected vertex it is deselected.

Place the  pointer inside a vertex, hold  down the <SHIFT> key, and press
the  *left* mouse button.   Release the button immediately without moving
the    mouse, release the   <SHIFT> key  afterwards.   If  the vertex was
deselected this action will  select it in  addition to any other selected
vertices. If the vertex was already selected it will be deselected.

Place the pointer  outside any vertex  and press the *left* mouse button.
Keep the button pressed and start moving the pointer.  This will create a
rubber band rectangle.  One corner at the position  where you pressed the
mouse button, the opposite corner following  the pointer.  As soon as you
release the left mouse button,   all  vertices inside the rectangle   are
selected, all vertices outside the rectangle are deselected.

Place the pointer outside any vertex, hold down the <SHIFT> key and press
the *left*  mouse button.  Again  you see a rubber  band, however, now as
soon as you release  the mouse button,  all vertices inside the rectangle
are selected in addition to any other selected vertices.

Call   the function 'Select' with   a subgroup or   list  of subgroups to
select.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{GraphicLattice, GAP Menu}

The <GAP> menu will  be pulled down if  you place the pointer inside  the
<GAP> button  and press the left mouse  button.  Keep the button down and
choose an entry by moving the pointer on  top of this entry.  Release the
mouse button to select an entry.

'save as postscript'

Selecting this entry pops up a file  selector.  If you  enter a file name
and click on <OK>, this will save a  description of the graphic sheet and
graphic objects on  the sheet as postscript  output.  However, only plain
postscript is created without any encapsulation whatsoever.

'close graphic sheet'

This entry will close the current graphic sheet and the window containing
the sheet  and all  <Information> menus related  to  this sheet.  On some
window system or using  some window manager the window  itself has also a
close button or menu entry.  However,  using the window close method will
not close associated <Information> menus.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{GraphicLattice, Resize Menu}

The <Resize> menu will be pulled down if you place the pointer inside the
<Resize> button and  press the left mouse  button.  Keep  the button down
and choose an entry by moving the pointer on top of  this entry.  Release
the mouse button to select an entry.

'Double Lattice'

Selecting the entry will  double the dimensions  of the graphic sheet and
enlarge the lattice accordingly.

'Halve Lattice'

Selecting  the entry will halve the  dimensions  of the graphic sheet and
shrink the lattice accordingly.

'Resize Lattice'

Selecting  this entry will  pop up a dialog   box asking for   an x and y
factor  separated by  a comma.   You   can enter integers,  quotients  or
decimal numbers. If you enter only  one number this is used  for x and y.
The graphic sheet is then enlarged or shrinked and the lattice is resized
accordingly.

'Resize Graphic Sheet'

This  entry is similar to 'Resize  Lattice' except that  only the graphic
sheet is changed, the lattice remains unchanged.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{GraphicLattice, Subgroups Menu}

The <Subgroups> menu will be pulled down  if you place the pointer inside
the <Subgroups> button and press the left mouse  button.  Keep the button
down and choose an entry   by moving the pointer  on  top of this  entry.
Release the mouse button to select an entry.

The result   of of a computation from   any of the  following  entries is
colored green, if your screen supports color.  There will also be a short
information message in the {\GAP} window about the result.

In the following descriptions   we use \"vertices\"\ as abbreviation  for
\"subgroups associated with vertices\".

'Centralizers'

computes  and displays the   centralizers of  the selected  vertices with
respect to the whole group.  Requires at  least one selected vertex.  See
also "Centralizer".

'Centres'

computes and displays  the centre of  the selected vertices.  Requires at
least one selected vertex. See also "Centre".

'Closure'

computes and    displays the common  closure  of  the  selected vertices.
Requires at least two selected vertices. See also "Closure".

'Closures'

computes  and  displays the pairwise  closures of  the selected vertices.
Requires at least three selected vertices. See also "Closure".

'Commutator Subgroups'

computes and displays  the pairwise commutator  subgroups of the selected
vertices.    Requires  at  least    two   selected  vertices. See    also
"CommutatorSubgroup".

'Cores'

computes and displays the cores of  the selected vertices with respect to
the whole group.  Requires at least one selected vertex. See also "Core".

'Derived Series'

For   each  selected vertex  'Derived  Series'  computes and displays its
derived series.   Requires  at   least  one selected vertex.    See  also
"DerivedSeries".

'Derived Subgroups'

computes and  displays the  derived  subgroups of the selected  vertices.
Requires at least one selected vertex.  See also "DerivedSubgroup".

'Fitting Subgroups'

computes and   displays   the   Fitting   subgroups   of   the   selected
vertices.  Requires    at   least one      selected  vertex.   See   also
"FittingSubgroup".

'Intersection'

computes and  displays the common intersection  of the selected vertices.
Requires at least two selected vertices.  See also "Intersection".

'Intersections'

computes and displays the pairwise intersection of the selected vertices.
Requires at least three selected vertices.  See also "Intersection".

'Normalizers'

computes   and displays  the normalizers   of  the selected vertices with
respect to the whole group.  Requires at least  one selected vertex.  See
also "Normalizer".

'Normal Closures'

For   each  selected vertex 'Normal  Closure'  computes  and displays the
normal closure of this vertex with respect  to the whole group.  Requires
at least one selected vertex.  See also "NormalClosure".

'Normal Subgroups'

For   each selected vertex 'Normal  Subgroups'  computes and displays the
normal subgroups of the subgroup associated  with this vertex.  These new
subgroups are not  necessarily  normal in the  whole  group.  Requires at
least one selected vertex.  See also "NormalSubgroups".

'Sylow Subgroups'

pops  up a dialog  box asking   for prime.   After entering  a  prime and
pressing  <return> or  clicking  <OK> it computes   and displays  a Sylow
subgroup  for   each selected vertex.  Requires   at   least one selected
vertex.  See also "SylowSubgroup".

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{GraphicLattice, CleanUp Menu}

The  <CleanUp> menu will  be pulled down if you  place the pointer inside
the <CleanUp> button and press  the left mouse  button.  Keep the  button
down and  choose an entry   by moving the pointer on   top of this entry.
Release the mouse button to select an entry.

'Average Y Levels'

selecting this entry will average the y coordinates of all vertices whose
associated subgroups have the same size.

'Average X Levels'

selecting this entry  will  average the x   coordinates  of two or   more
selected vertices.  This will only work if the corresponding subgroups do
not have the same size. Requires at least two selected vertices.

'Rotate Conjugates'

rotates the selected vertices of a conjugacy class cyclicly.  Requires at
least two selected vertices.

'Deselect All'

deselects all vertices.  Requires at least one selected vertex.

'Relabel Vertices'

selecting  this entry will pop  up a dialog  box for each selected vertex
asking for a new label.  Clicking on <CANCEL>  will cancel the relabeling
of the remaining vertices but will not reset the already changed labels.

'Select Representatives'

deselects  all vertices    and  then selects a   representative  for each
conjugacy class.

'Use Black\&White'

switches to black and white in case of a color screen.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{GraphicLattice, Information Menu}

Placing the pointer  inside a vertex (selected  or not)  and pressing the
*right* mouse button pops up the  <Information> menu.  Clicking on any of
the text  lines will compute  the corresponding  property of the subgroup
$u$ associated  with  this vertex.  Clicking   on <all> will  compute all
properties, clicking on <close> will close the <Information> menu.

'Size'

computes and displays the size of $u$.  See also "Size".

'Index'

computes and displays the  index  of $u$ in   the whole group.  See  also
"Index".

'IsAbelian' \\
'IsCyclic' \\
'IsNilpotent' \\
'IsPerfect' \\
'IsSimple' \\
'IsSolvable'

computes  and displays  the corresponding property    of $u$.   See  also
"IsAbelian",  "IsCyclic",  "IsNilpotent",  "IsPerfect",  "IsSimple",  and
"IsSolvable".


'IsCentral' \\
'IsNormal'

computes and displays the corresponding  property of $u$ with respect  to
the whole group.  See also "IsCentral" and "IsNormal".

'Isomorphism'

computes and displays  the isomorphism type of  $u$.  This will only work
if the size of $u$ is small.  See "GroupId" for details.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{GraphicLattice, Sheet Menu}

In order to adjust any margin place the pointer outside of any vertex and
press the  *right* mouse  button,  keep the button  pressed.  The <Sheet>
menu will pop up.

'Left Margin'

selecting <Left  Margin> will narrow the  graphic sheet, setting  the new
left margin to the x coordinate of your mouse click.

'Right Margin'

selecting <Right  Margin> will narrow the graphic  sheet, setting the new
right margin to the x coordinate of your mouse click.

'Top Margin'

selecting <Top Margin> will narrow the graphic sheet, setting the new top
margin to the y coordinate of your mouse click.

'Bottom Margin'

selecting <Bottom Margin> will narrow the graphic  sheet, setting the new
bottom margin to the y coordinate of your mouse click.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{InteractiveLattice}

'InteractiveLattice( <g> )'

'InteractiveLattice' creates a new graphic sheet $s$ containing the Hasse
diagram of a partial subgroup lattice of a group <g>.  Initially only two
vertices representing <g> and the trivial subgroup are present.

Note that    <g>  must  not   be a    finitely  presented  group.     See
"InteractiveLattice for FpGroups" if <g> is a finitely presented group.

Moving  vertices, selecting vertices,  the <GAP>, <Resize>, <Information>
and <sheet> menus   are  identical to  the  'GraphicLattice'  case.   See
"GraphicLattice,  Moving Vertices", "GraphicLattice, Selecting Vertices",
"GraphicLattice,    GAP       Menu",    "GraphicLattice,  Resize   Menu",
"GraphicLattice,  Information Menu",  and  "GraphicLattice,  Sheet  Menu"
respectively.

The next  next sections describe the  differences in  the <Subgroups> and
<CleanUp> menus and the different vertex shapes.

The inclusions shown  in the Hasse  diagram of  the  partial lattice will
always be correct.  New subgroups are always  be checked for conjugation,
if  the  normalizer  of   the a  new   subgroup  has an index   less than
'$s$.limits.conjugates'.  If this number is negative the check is carried
out     if   the     index of   the       new  subgroup   is   less  than
'-$s$.limits.conjugates'.

'InteractiveLattice( <g>, <width>, <height> )'

In  this  form 'InteractiveLattice' creates   a  graphic sheet of initial
dimensions <width>  times <height>.   It   is however still possible   to
change these dimensions  later using either "GraphicLattice, Resize Menu"
or "GraphicLattice, Sheet Menu".

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{InteractiveLattice, Subgroups Menu}

In addition to "GraphicLattice, Subgroups Menu" the following entries are
available.

'Conjugate Subgroups'

computes and displays the conjugacy  classes (with  respect to the  whole
group) of each  selected vertex.  Requires  at least one selected vertex.
See also "ConjugacyClass".

'Random Conjugates'

computes and  displays   a  random conjugate  for  each  conjugacy  class
containing a selected vertex.  Requires at least one selected vertex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{InteractiveLattice, CleanUp Menu}

In addition  to "GraphicLattice, Subgroups  Menu"  the following entry is
available.

'Merge Classes'

merges two conjugacy classes. Use this in  case the conjugation check was
not  carried out   (see "InteractiveLattice")  but you  know   that a new
subgroup is conjugate to an old one.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{InteractiveLattice, Vertex Shapes}

The following vertex shapes can appear in an interactive lattice.

circle:\\
    conjugation for the subgroup associated with  the vertex was checked.
    If the subgroup is conjugate  to any other  subgroup of a vertex both
    vertices are drawn closely together.

square:\\
    conjugation was not  checked  for the  subgroup associated with  this
    vertex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{InteractiveLattice for FpGroups}

'InteractiveLattice( <g> )'

'InteractiveLattice' creates a new graphic sheet $s$ containing the Hasse
diagram of a  partial subgroup lattice of  finitely presented  group <g>.
Initially only one vertex representing <g> is present.

See "InteractiveLattice" if <g> is not a finitely presented group.

Moving  vertices, selecting  vertices,  the <GAP>, <Resize>,  and <sheet>
menus are identical  to the 'GraphicLattice' case.   See "GraphicLattice,
Moving  Vertices", "GraphicLattice, Selecting Vertices", "GraphicLattice,
GAP Menu",  "GraphicLattice,   Resize Menu", and  "GraphicLattice,  Sheet
Menu" respectively.

'InteractiveLattice( <g>, <width>, <height> )'

In  this  form 'InteractiveLattice' creates   a  graphic sheet of initial
dimensions <width>  times <height>.   It   is however still possible   to
change these dimensions  later using either "GraphicLattice, Resize Menu"
or "GraphicLattice, Sheet Menu".

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{InteractiveLattice for FpGroups, Subgroups Menu}

The <Subgroups> menu will be pulled down  if you place the pointer inside
the <Subgroups> button and press the  left mouse button.  Keep the button
down and  choose an entry by  moving  the pointer on   top of this entry.
Release the mouse button to select an entry.

The result of  of  a computation  from any  of  the following entries  is
colored green, if  your screen supports color.   In most cases there will
also be short information message in the {\GAP} window about the result.

In  the following descriptions, we  use \"vertices\"\ as abbreviation for
\"subgroups associated with vertices\".

'Abelian Prime Quotient'

pops  up a dialog  box asking  for a  prime  $p$.  It  then computes  and
displays the largest elementary   abelian  $p$ quotient of  the  selected
vertex.  If no presentation for the  subgroup associated to the vertex is
known   a presentation is  first computed  using  a modified Todd-Coxeter
algorithm.     It then calls    'PrimeQuotient' to   compute  the largest
elementary abelian quotient.  Requires exactly one selected vertex.

'All Overgroups'

computes and displays all over-groups  of the selected vertex.  It  first
computes the permutation action of the whole  group on the consets of the
subgroup associated with  the selected vertex and  then searches  for all
block systems.  Requires exactly one selected vertex.

'Conjugacy Class'

computes and  displays  the   conjugacy  class of  the   selected vertex.
Requires  exactly  one selected vertex.  

'Cores'

computes and  displays the cores  of the selected vertices.   Requires at
least one selected vertex.


'Derived Subgroups'

computes and displays the derived subgroups of the selected vertices.  It
will  only  display  those derived    subgroups  whose index is   finite.
Requires at least one selected vertex.

'Epimorphisms'

pops up another menu. Requires exactly one selected vertex.

|    Sym(n)
    Alt(n)
    Sym(n) > Alt(n)
    PSL(d,q)
    Library
    User Defined |

Click on any of these entries to try to find a quotient isomorphic to the
symmetric  group (|Sym(n)|), the     alternating group  (|Alt(n)|),   the
symmetric or the alternating   group (|Sym(n) > Alt(n)|),  the projective
special  linear group  (|PSL(d,q)|), a group  in a  library supplied with
{\XGAP} (this will   pops up a file   selector), or a user defined  group
stored   in   the  variable  |PERM_GROUP|.   After  supplying  additional
parameters,  for  example, the  degree of   the   symmetric group  or the
dimension and field of $PSL$  using dialog boxes, the corresponding entry
will change, for example to something like

|    Sym(3)        3 found|

After one or more quotients were found click <display> to display them.

'Intersections'

computes and   displays   the pairwise   intersections  of  the  selected
vertices.  Requires at least two selected vertices.

'Low Index Subgroups'

pops up a dialog box asking for index  limit $k$.  It  will then do a low
index subgroup search for  subgroup of index at  most $k$ of the selected
vertex using 'LowIndexSubgroupsFpGroup'.    If no  presentation  for  the
subgroup   associated to  the vertex  is  known a  presentation  is first
computed using a  modified  Todd-Coxeter algorithm. Requires  exactly one
selected vertex.

'Normalizers'

computes and displays the normalizers of the selected vertices.  Requires
at least one selected vertex.

'Prime Quotient'

pops up a dialog box asking for a prime $p$ and another dialog box asking
for a class $c$.  It then  computes and displays the largest $p$-quotient
of class $c$ of the selected vertex.  If no presentation for the subgroup
associated to the vertex is known a  presentation is first computed using
a   modified  Todd-Coxeter algorithm.   It    then calls 'PrimeQuotient'.
Requires exactly one selected vertex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{InteractiveLattice for FpGroups, CleanUp Menu}

'Check Coset Table'

will compute  coset tables  for the selected   vertices and  checks these
coset  tables against the other known  tables.  If equal coset tables are
discovered the corresponding vertices are  merged.  Requires at least one
selected vertex.

'Deselect All'

deselects all vertices.  Requires at least one selected vertex.

'Delete Vertex'

deletes  a selected vertex.   *Handle with  care*, if  the vertex is  the
start  point  of  a  prime quotient  or   a derived subgroup all  related
vertices are deleted.  Requires at least one selected vertex.

'Use Black\&White'

switches to black and white in case of a color screen.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{InteractiveLattice for FpGroups, Information Menu}

Placing the pointer  inside a vertex (selected  or not) and  pressing the
*right* mouse button pops up the <Information>  menu.  Clicking on any of
the text  lines will compute the corresponding   property of the subgroup
$u$  associated  with this  vertex. Clicking  on <close>  will  close the
<Information> menu.  Let $s$ be the graphic sheet.

'Index'

displays  the index of  $u$ in the whole  group.  The  index is stored in
'$s$.result'.

'IsNormal'

checks  if $u$ is  normal  in the  whole group.    The result  ('true' or
'false') is stored in '$s$.result'.

'Abelian Invariants'

computes  and  displays  the  abelian invariants  of  $u$.   The  abelian
invariants are stored in '$s$.result'.

'Presentation'

computes a presentation of $u$ and displays  the number of generators and
relators   of   this presentation.     The   presentation is    stored in
'$s$.result'.

'Coset Table'

computes   a   coset table for   $u$.    The  coset  table  is  stored in
'$s$.result'.

'Factor Fp Group'

computes a presentation  for the factor group  of the whole group by $u$,
if $u$ is normal.  The presentation is stored in '$s$.result'.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{InteractiveLattice for FpGroups, Vertex Shapes}

The following vertex shapes can appear in an interactive lattice.

circle:\\
    a vertex for which  a coset table for  associated subgroup is  known.
    These vertices are checked for equality among each other.

square:\\
    a vertex for which  no coset table for  associated subgroup is known.
    These vertices are not checked for equality among all vertices.

diamond:\\
    as  a 'circle' but   the associated subgroup  is  normal in the whole
    group *and* {\GAP} knows this.

diamond with square:\\
    as a  'square'  but the associated subgroup  is  normal in  the whole
    group *and* {\GAP} knows this.
