%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%A  copyrigh.tex                GAP documentation            Martin Schoenert
%%
%A  @(#)$Id: copyrigh.tex,v 1.1.1.1 1996/12/11 12:36:44 werner Exp $
%%
%Y  Copyright 1990-1992,  Lehrstuhl D fuer Mathematik,  RWTH Aachen,  Germany
%%
%%  This file contains the GAP copyright
%%
%H  $Log: copyrigh.tex,v $
%H  Revision 1.1.1.1  1996/12/11 12:36:44  werner
%H  Preparing 3.4.4 for release
%H
%H  Revision 3.3  1992/04/05  23:41:40  martin
%H  made a minor modification
%H
%H  Revision 3.2  1992/03/31  12:50:36  martin
%H  fixed several typos
%H
%H  Revision 3.1  1992/03/31  08:22:32  martin
%H  initial revision under RCS
%H
%%
\thispagestyle{empty}

{\large Copyright {\copyright} 1992 by Lehrstuhl D f{\accent127u}r Mathematik}

RWTH, Templergraben 64, D 5100 Aachen, Germany

{\GAP}  can  be  copied  and  distributed  freely for any  non-commercial
purpose.

If you copy {\GAP} for somebody else, you may ask this person for  refund
of your expenses.  This should cover cost of media, copying and shipping.
You are not allowed to ask for more than this.  In any case you must give
a copy of this copyright notice along with the program.

If you obtain {\GAP} please send us  a short notice to that effect, e.g.,
an  e-mail  message   to  the  address  'gap@samson.math.rwth-aachen.de',
containing your full  name and address.  This  allows us to keep track of
the number of {\GAP} users.

If you  publish  a mathematical result  that  was  partly obtained  using
{\GAP}, please cite {\GAP}, just as you would cite another paper that you
used.   Also we  would appreciate it if you could inform us about  such a
paper.

You  are permitted  to modify and  redistribute  {\GAP},  but you are not
allowed  to restrict further redistribution.  That is to say  proprietary
modifications will  not  be allowed.  We want all  versions  of {\GAP} to
remain free.

If you  modify any part of {\GAP} and redistribute it,  you must supply a
'README'  document.   This should specify what modifications you made  in
which  files.  We do  not  want  to take  credit  or  be blamed  for your
modifications.

Of course we are interested in all of your modifications.  In  particular
we would like to see bug-fixes, improvements and new functions.  So again
we would appreciate it if you would inform us about all modifications you
make.

{\GAP} is distributed by us without any warranty, to the extent permitted
by applicable state law.  We  distribute {\GAP} *as is* without  warranty
of any kind, either expressed or implied, including,  but not limited to,
the implied  warranties  of merchantability and  fitness for a particular
purpose.

The entire risk as to the quality and performance of the program is  with
you.  Should {\GAP} prove defective, you assume the cost of all necessary
servicing, repair or correction.

In no  case  unless  required by applicable law will we, and/or any other
party who  may  modify  and  redistribute  {\GAP}  as permitted above, be
liable  to you for damages, including lost profits, lost monies or  other
special, incidental or consequential damages  arising out  of the  use or
inability to use {\GAP}.
