%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%A  language.tex                GAP documentation            Martin Schoenert
%%
%A  @(#)$Id: language.tex,v 1.1.1.1 1996/12/11 12:36:47 werner Exp $
%%
%Y  Copyright 1990-1992,  Lehrstuhl D fuer Mathematik,  RWTH Aachen,  Germany
%%
%%  This file describes the {\GAP} programming language.
%%
%H  $Log: language.tex,v $
%H  Revision 1.1.1.1  1996/12/11 12:36:47  werner
%H  Preparing 3.4.4 for release
%H
%H  Revision 3.11  1993/03/11  17:54:26  fceller
%H  strings are now lists
%H
%H  Revision 3.10  1993/02/19  10:48:42  gap
%H  adjustments in line length and spelling
%H
%H  Revision 3.9  1993/02/18  15:52:09  felsch
%H  index entries added
%H
%H  Revision 3.8  1993/02/17  07:43:50  felsch
%H  examples fixed
%H
%H  Revision 3.7  1993/02/11  12:20:47  martin
%H  changed reference "Strings" to "Strings and Characters"
%H
%H  Revision 3.6  1993/02/10  21:10:03  martin
%H  fixed various typos, extended description for 3.2
%H
%H  Revision 3.5  1992/04/30  12:07:10  martin
%H  changed a few sentences to avoid bold non-roman fonts
%H
%H  Revision 3.4  1992/04/09  11:36:01  martin
%H  made a few changes so that two LaTeX passes suffice
%H
%H  Revision 3.3  1992/04/02  20:58:24  martin
%H  fixed a wrong example in "While"
%H
%H  Revision 3.2  1992/03/11  15:59:22  martin
%H  fixed several typos
%H
%H  Revision 3.1  1992/01/14  14:42:24  martin
%H  changed the BNF for nicer online viewing
%H
%H  Revision 3.0  1991/12/27  16:10:27  martin
%H  initial revision under RCS
%H
%%
\Chapter{The Programming Language}

This chapter describes the {\GAP} programming language.   It should allow
you in principle to predict the result of each and every input.  In order
to know what we are talking about, we first have to  look more closely at
the process of  interpretation  and the  various representations  of data
involved.

First we have the input to {\GAP}, given as a string  of characters.  How
those characters enter {\GAP} is operating system  dependent,  e.g., they
might be entered at a  terminal, pasted with  a mouse into  a window,  or
read from a file.  The mechanism does not matter.  This representation of
expressions by characters is called the *external representation* of  the
expression.  Every expression has  at  least one external  representation
that can be entered to get exactly this expression.

The input, i.e., the external representation, is transformed in a process
called *reading* to an internal representation.   At this point the input
is analyzed  and inputs   that  are  not legal external  representations,
according to the rules given below, are rejected  as errors.  Those rules
are usually called the *syntax* of a programming language.

The  internal  representation  created by  reading  is called  either  an
*expression* or a *statement*.   Later we  will distinguish between those
two terms, however now we will use them interchangeably.  The exact  form
of the internal  representation does not matter.  It could be a string of
characters  equal  to  the external  representation,  in  which case  the
reading  would  only  need to check for errors.  It could be a series  of
machine instructions  for the  processor  on  which {\GAP} is running, in
which case the reading would  more appropriately  be  called compilation.
It is in fact a tree--like structure.

After the input has been read it is again transformed in a process called
*evaluation* or *execution*.  Later we will distinguish between those two
terms too, but for the moment we will use them interchangeably.  The name
hints at the nature of this  process, it replaces an  expression with the
value of the expression.  This  works  recursively,  i.e., to evaluate an
expression  first the subexpressions  are evaluated and then the value of
the expression is  computed according  to rules  given  below from  those
values.  Those rules  are usually called the *semantics* of a programming
language.

The result of the evaluation is, not surprisingly, called a *value*.  The
set of values    is  of course a    much smaller  set  than  the   set of
expressions;  for  every value there  are   several expressions that will
evaluate to   this  value.  Again  the form  in  which such  a   value is
represented  internally does not   matter.  It is    in fact a tree--like
structure again.

The last process  is called *printing*.  It  takes the  value produced by
the evaluation and creates an external representation,  i.e., a string of
characters again.  What you do with this external representation is up to
you. You can look at it, paste it with the  mouse into another window, or
write it to a file.

Lets look at an example to make this more clear.  Suppose you type in the
following string of 8 characters

|    1 + 2 * 3;|

{\GAP}  takes   this external representation   and creates   a  tree like
internal representation, which we can picture as follows

|       +
      / \
     1   *
        / \
       2   3 |

This expression is then evaluated.  To do this {\GAP} first evaluates the
right subexpression '2\*3'.   Again to do this {\GAP} first evaluates its
subexpressions 2  and 3.  However they  are so simple that they are their
own value, we say that  they are  self--evaluating.  After  this has been
done,  the rule for '\*' tells us that the value is the  product  of  the
values of the  two  subexpressions,  which  in  this case  is clearly  6.
Combining  this with the value  of the left operand of  the '+', which is
self--evaluating too gives us  the value of the whole expression 7.  This
is  then  printed,  i.e.,  converted  into  the  external  representation
consisting of the single character '7'.

In this fashion  we can predict the  result of every input  when we know
the syntactic rules that govern the  process of reading and the semantic
rules that  tell us for  every expression how  its value is  computed in
terms of the values of the subexpressions. The syntactic rules are given
in  sections  "Lexical  Structure", "Language  Symbols",  "Whitespaces",
"Keywords", "Identifiers", and  "The Syntax in BNF",  the semantic rules
are  given in  sections  "Expressions",  "Variables", "Function  Calls",
"Comparisons",  "Operations",  "Statements",  "Assignments",  "Procedure
Calls", "If",  "While", "Repeat",  "For", "Functions", and  the chapters
describing the individual data types.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Lexical Structure}

The input of {\GAP} consists of sequences of the following characters.

Digits, uppercase and lowercase letters,  <space>, <tab>, <newline>,  and
the special characters

|    "       '       (       )       *       +       ,       _
    .       /       :       ;       <       =       >       ~
    [       \       ]       ^       _       {       }       & |

Other  characters  will  be signalled  as  illegal.   Inside strings  and
comments the full character set supported by the computer is allowed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Language Symbols}

The  process of reading,  i.e., of assembling the input into expressions,
has a subprocess,  called  *scanning*, that assembles the characters into
symbols.   A *symbol*  is  a sequence of characters that form  a  lexical
unit.  The  set of  symbols  consists of  keywords, identifiers, strings,
integers, and operator and delimiter symbols.

A keyword is a  reserved word  consisting entirely  of lowercase  letters
(see "Keywords").  An identifier is a sequence of letters and digits that
contains at  least one letter  and is not a keyword  (see "Identifiers").
An integer is a  sequence of  digits  (see  "Integers").   A string is  a
sequence  of   arbitrary  characters   enclosed  in  double  quotes  (see
"Strings and Characters").

Operator and delimiter symbols are

|    +       -       *       /       ^       ~
    =       <>      <       <=      >       >=
    :=      .       ..      ->      ,       ;
    [       ]       {       }       (       )  |

Note  that during the process of  scanning also all whitespace is removed
(see "Whitespaces").

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Whitespaces}%
\index{space}%
\index{blank}\index{tabulator}\index{newline}\index{comments}

The    characters <space>,  <tab>,   <newline>, and   <return> are called
*whitespace characters*.   Whitespace is used   as necessary to  separate
lexical symbols, such as integers, identifiers, or keywords.  For example
'Thorondor' is  a single identifier, while 'Th  or  ondor' is the keyword
'or' between the two identifiers 'Th'  and 'ondor'.  Whitespace may occur
between any two  symbols, but not within a  symbol.  Two or more adjacent
whitespaces are  equivalent to a single  whitespace.  Apart from the role
as  separator  of  symbols,  whitespaces  are   otherwise  insignificant.
Whitespaces may also  occur inside a string,  where they are significant.
Whitespaces should also be used freely for improved readability.

A  *comment* starts with the   character '\#', which is sometimes  called
sharp or hatch, and continues to the end of the line on which the comment
character appears.  The whole  comment, including '\#' and  the <newline>
character  is  treated  as a single    whitespace.  Inside a  string, the
comment character '\#' looses its role and is just an ordinary character.

For example, the following statement

|    if i<0 then a:=-i;else a:=i;fi; |

is equivalent to

|    if i < 0  then      & if i is negative
        a := -i;        &     take its inverse
    else                & otherwise
        a := i;         &     take itself
    fi; |

(which by the  way  shows  that  it  is  possible  to  write  superfluous
comments).  However the first statement is not equivalent to

|    ifi<0thena:=-i;elsea:=i;fi; |

since  the keyword 'if'  must  be separated from  the identifier 'i' by a
whitespace,  and  similarly 'then'  and 'a',  and 'else' and  'a' must be
separated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Keywords}

*Keywords* are reserved words that are used  to denote special operations
or  are part of  statements.  They must not be used  as identifiers.  The
keywords are

|    and       do        elif      else      end       fi
    for       function  if        in        local     mod
    not       od        or        repeat    return    then
    until     while     quit |

Note that all keywords are written in lowercase.  For example only 'else'
is   a  keyword;  'Else', 'eLsE',  'ELSE'   and   so  forth  are ordinary
identifiers.  Keywords must not contain  whitespace, for  example 'el if'
is not the same as 'elif'.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Identifiers}

An identifier is used to  refer to   a  variable (see  "Variables").   An
identifier consists of letters,  digits,  and underscores '\_',  and must
contain at least one letter  or underscore.   An identifier is terminated
by the first character not in this class.  Examples  of valid identifiers
are

|    a                   foo                 aLongIdentifier
    hello               Hello               HELLO
    x100                100x                _100
    some_people_prefer_underscores_to_separate_words
    WePreferMixedCaseToSeparateWords |

Note that  case  is  significant, so  the three identifiers in the second
line are distinguished.

The backslash |\| can be used to include other characters in identifiers;
a backslash  followed  by  a character  is equivalent   to the character,
except that this escape sequence  is considered to be an ordinary letter.
For example |G\(2\,5\)| is an identifier, not a call to a function 'G'.

An identifier that  starts with  a backslash is  never a  keyword, so for
example |\*| and |\mod| are identifier.

The length  of identifiers is not limited,   however only  the first 1023
characters are significant.  The escape sequence |\|<newline> is ignored,
making it possible to split long identifiers over multiple lines.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Expressions}%
\index{evaluation}

An  *expression* is a construct  that  evaluates  to a value.   Syntactic
constructs that are executed to produce a side effect and return no value
are called *statements*  (see "Statements").  Expressions appear as right
hand  sides of assignments  (see "Assignments"),  as  actual arguments in
function calls (see "Function Calls"), and in statements.

Note that an expression is not the same as a value.  For example '1 + 11'
is  an    expression, whose value is   the   integer  12.    The external
representation of this integer is the character sequence '12', i.e., this
sequence is output  if the integer is  printed.  This sequence is another
expression whose value is  the integer 12.    The process of finding  the
value of an  expression  is done by  the  interpreter and is  called  the
*evaluation* of the expression.

Variables,  function  calls, and integer, permutation, string,  function,
list, and record literals (see "Variables", "Function Calls", "Integers",
"Permutations",   "Strings   and   Characters",   "Functions",   "Lists",
"Records"), are the simplest cases of expressions.

Expressions,  for example the simple expressions  mentioned above, can be
combined with the operators to form  more complex expressions.  Of course
those expressions can then be combined further with the operators to form
even more complex  expressions.  The *operators* fall into three classes.
The  *comparisons*  are  '=', '\<>',  '\<=', '>',   '>=',  and  'in' (see
"Comparisons" and "In").  The *arithmetic operators* are  '+', '-', '\*',
'/', 'mod', and '\^'   (see "Operations").   The *logical  operators* are
'not', 'and', and 'or' (see "Operations for Booleans").

|    gap> 2 * 2;    # a very simple expression with value
    4
    gap> 2 * 2 + 9 = Fibonacci(7) and  Fibonacci(13) in Primes;
    true            # a more complex expression |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Variables}%
\index{scope}\index{bound}

A *variable*  is a location in a  {\GAP} program that  points to a value.
We say the variable is *bound* to this value.  If a variable is evaluated
it evaluates to this value.

Initially an ordinary variable is not  bound to  any value.  The variable
can be bound to a  value  by *assigning* this value to  the variable (see
"Assignments").  Because of this we sometimes say that a variable that is
not bound to any value has no assigned value.  Assignment  is in fact the
only way by which a variable, which is not an argument of a function, can
be bound to  a  value.  After a variable  has  been bound to a   value an
assignment can also be used to bind the variable to another value.

A special class of  variables are *arguments* of functions.  They  behave
similarly to other variables, except they are  bound to the  value of the
actual arguments upon a function call (see "Function Calls").

Each variable has a  name that is also called  its *identifier*.  This is
because in  a given scope an identifier identifies a unique variable (see
"Identifiers").  A *scope* is a lexical part of a program text.  There is
the  global scope that encloses the  entire program  text, and there  are
local  scopes that  range  from  the  'function'  keyword,  denoting  the
beginning of a function  definition, to  the corresponding 'end' keyword.
A local scope  introduces new  variables, whose identifiers are given  in
the formal argument list and the 'local' declaration of the function (see
"Functions").  Usage  of an  identifier in  a program  text refers to the
variable in  the innermost scope  that has  this identifier as  its name.
Because  this  mapping  from  identifiers  to variables is done  when the
program is read, not when it is executed, {\GAP} is  said to have lexical
scoping.   The  following example  shows  how  one identifier  refers  to
different variables at different points in the program text.

|     g := 0;            & global variable g
     x := function ( a, b, c )
        local   y;
        g := c;         & c refers to argument c of function x
        y := function ( y )
            local  d, e, f;
            d := y;     & y refers to argument y of function y
            e := b;     & b refers to argument b of function x
            f := g;     & g refers to global variable g
            return d + e + f;
        end;
        return y( a );  & y refers to local y of function x
    end; |

It is important to note that the concept of a variable in {\GAP} is quite
different from the concept  of  a variable in programming languages  like
PASCAL.  In  those languages a variable  denotes a block of  memory.  The
value of the variable is stored in this block.  So in those languages two
variables can  have the  same value,  but  they can never  have identical
values,  because  they  denote different  blocks of  memory.   (Note that
PASCAL has  the concept of a reference  argument.  It seems as if such an
argument and the variable used in the actual function  call have the same
value, since changing the argument\'s value also changes the value of the
variable  used in  the actual  function call.   But this is  not  so; the
reference  argument is actually  a pointer  to the variable  used  in the
actual function call, and it is the compiler that inserts enough magic to
make the  pointer  invisible.)  In  order  for this to work the  compiler
needs enough  information to compute the amount of memory needed for each
variable in a  program,  which is  readily  available in the declarations
PASCAL requires for every variable.

In {\GAP} on the other hand each variable justs points to a value.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Function Calls}

'<function-var>()' \\
'<function-var>( <arg-expr> \{',' <arg-expr>\} )'

The function call has the effect  of calling the function <function-var>.
The precise semantics are as follows.

First  {\GAP} evaluates the <function-var>.  Usually  <function-var> is a
variable,  and  {\GAP} does nothing more  than taking  the value of  this
variable.   It is allowed  though  that  <function-var> is a more complex
expression,  namely it can for example be  a selection of  a list element
'<list-var>[<int-expr>]',  or   a   selection  of   a   record  component
'<record-var>.<ident>'.  In any  case {\GAP} tests whether the value is a
function.  If it is not, {\GAP} signals an error.

Next {\GAP} checks that the number of actual arguments <arg-expr>s agrees
with the number of formal arguments as given  in the function definition.
If they do not agree {\GAP}  signals an error.  An  exception is the case
when there is  exactly one formal argument  with the name 'arg', in which
case any number of actual arguments is allowed.

Now {\GAP} allocates for each formal argument and for each formal local a
new variable.  Remember that a variable is a location in a {\GAP} program
that points  to  a value.  Thus for   each formal argument  and  for each
formal local such a location is allocated.

Next the arguments <arg-expr>s are evaluated, and the values are assigned
to the newly created variables corresponding to the formal arguments.  Of
course the first value  is assigned  to the new variable corresponding to
the  first  formal argument, the second   value   is  assigned to the new
variable corresponding   to   the second    formal argument, and   so on.
However, {\GAP} does not make any guarantee about the order  in which the
arguments are evaluated.  They might be evaluated left to right, right to
left, or in  any  other order, but  each  argument is evaluated once.  An
exception again occurs if the function has  only one formal argument with
the name 'arg'.  In this case the values  of all the actual arguments are
stored  in   a  list and this   list  is assigned  to the    new variable
corresponding to the formal argument 'arg'.

The  new variables corresponding to the  formal  locals are initially not
bound to any   value.   So trying   to evaluate  those   variables before
something has been assigned to them will signal an error.

Now the body of the function, which is  a statement, is executed.  If the
identifier of one of the formal arguments or formal locals appears in the
body of the function it refers to the new variable that was allocated for
this formal argument or formal local, and evaluates to  the value of this
variable.

If during the execution of the body of the function  a 'return' statement
with  an expression (see "Return") is executed,  execution of the body is
terminated  and  the value  of  the function  call  is the value  of  the
expression  of  the 'return'.  If  during the  execution of  the  body  a
'return'  statement  without an expression is executed,  execution of the
body  is  terminated and the function  call does not produce  a value, in
which case  we call this call a procedure  call (see  "Procedure Calls").
If the  execution of  the body completes without execution of  a 'return'
statement, the  function call again produces no  value, and again we talk
about a procedure call.

|    gap> Fibonacci( 11 );
        # a call to the function 'Fibonacci' with actual argument '11'
    89 |
        
|    gap> G.operations.RightCosets( G, Intersection( U, V ) );;
        # a call to the function in 'G.operations.RightCosets'
        # where the second actual argument is another function call |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Comparisons}

'<left-expr> =   <right-expr>' \\
'<left-expr> \<> <right-expr>'

The operator '=' tests for equality of  its two operands and evaluates to
'true' if they are equal and to 'false'  otherwise.  Likewise '\<>' tests
for  inequality of its  two operands.  Note  that any two  objects can be
compared, i.e., '=' and '\<>' will never  signal an error.  For each type
of objects the definition of equality is given in the respective chapter.
Objects of different types are  never equal, i.e.,  '=' evaluates in this
case to 'false', and '\<>' evaluates to 'true'.

'<left-expr> \<\ <right-expr>' \\
'<left-expr> >   <right-expr>' \\
'<left-expr> \<= <right-expr>' \\
'<left-expr> >=  <right-expr>'

'\<' denotes less than,  '\<=' less than or  equal, '>' greater than, and
'>=' greater than or equal of its two operands.  For each type of objects
the definition of the  ordering is given in  the respective chapter.  The
ordering  of  objects  of different types  is as follows.  Rationals  are
smallest,  next  are  cyclotomics, followed  by  finite  field  elements,
permutations, words, words in solvable groups, boolean values, functions,
lists, and records are largest.

Comparison operators, which includes the operator 'in' (see "In") are not
associative, i.e., it is not allowed to write '<a> =  <b> \<> <c> = <d>',
you  must use '(<a>  =  <b>) \<>  (<c>  = <d>)'  instead.  The comparison
operators have   higher precedence   than    the logical  operators  (see
"Operations  for  Booleans"),  but lower  precedence  than the arithmetic
operators (see "Operations").  Thus, for example, '<a> \*\ <b> =  <c> and
<d>' is interpreted, '((<a> \*\ <b>) = <c>) and <d>)'.

|    gap> 2 * 2 + 9 = Fibonacci(7);    # a comparison where the left
    true                              # operand is an expression |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Operations}%
\index{precedence}\index{associativity}

'+ <right-expr>' \\
'- <right-expr>' \\
'<left-expr> +   <right-expr>' \\
'<left-expr> -   <right-expr>' \\
'<left-expr> \*\ <right-expr>' \\
'<left-expr> /   <right-expr>' \\
'<left-expr> mod <right-expr>' \\
'<left-expr> \^\ <right-expr>'

The arithmetic operators are  '+',  '-',  '\*',  '/',  'mod',  and  '\^'.
The meanings (semantic) of those  operators generally depend on the types
of the operands  involved, and they are  defined in the  various chapters
describing the types.  However basically the meanings are as follows.

'+' denotes the addition,  and '-'  the  subtraction  of ring  and  field
elements.   '\*' is  the  multiplication of  group  elements,  '/' is the
multiplication of the left operand with the inverse of the right operand.
'mod' is only defined for integers  and rationals and denotes  the modulo
operation.  '+' and '-' can also be used as unary  operations.  The unary
'+' is ignored and unary '-' is equivalent to multiplication by -1.  '\^'
denotes powering of a group  element  if the right operand is an integer,
and is  also used to denote operation if  the  right operand  is  a group
element.

The *precedence* of those operators is as follows.  The powering operator
'\^' has  the highest precedence, followed by the unary operators '+' and
'-',  which are followed by the  multiplicative  operators '\*', '/', and
'mod', and  the additive  binary operators '+' and  '-'  have  the lowest
precedence.   That  means that the expression  '-2 \^\ -2 \*\  3 + 1'  is
interpreted as '(-(2  \^\ (-2)) \*\ 3) + 1'.  If in doubt use parentheses
to clarify your intention.

The *associativity* of the arithmetic operators is as follows.'\^' is not
associative,  i.e., it is  illegal to write '2\^3\^4', use parentheses to
clarify whether  you mean '(2\^3)  \^\  4' or '2 \^\ (3\^4)'.  The  unary
operators '+' and '-' are right associative, because they are written  to
the left of their operands.  '\*', '/', 'mod', '+',  and '-' are all left
associative, i.e., '1-2-3' is interpreted as '(1-2)-3'  not as '1-(2-3)'.
Again, if in doubt use parentheses to clarify your intentions.

The  arithmetic  operators  have higher  precedence  than  the comparison
operators (see  "Comparisons" and  "In")  and the logical  operators (see
"Operations for  Booleans").  Thus, for  example, '<a> \*\ <b> =  <c> and
<d>' is interpreted, '((<a> \*\ <b>) = <c>) and <d>'.

|    gap> 2 * 2 + 9;    & a very simple arithmetic expression
    13 |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Statements}%
\index{execution}

Assignments (see "Assignments"), Procedure calls (see "Procedure Calls"),
'if'  statements  (see  "If"),  'while'  (see  "While"),  'repeat'   (see
"Repeat") and  'for'  loops (see "For"), and the 'return'  statement (see
"Return") are called statements.  They can be entered interactively or be
part of a function definition.  Every statement must  be terminated by  a
semicolon.

Statements, unlike expressions, have no value.  They are executed only to
produce an effect.  For example an assignment has the effect of assigning
a   value to a  variable, a  'for' loop   has the  effect  of executing a
statement sequence for  all elements in a  list and so  on.  We will talk
about *evaluation* of expressions  but about *execution* of statements to
emphasize this difference.

It is possible to use expressions as statements.  However this does cause
a warning.

|    gap> if i <> 0  then  k = 16/i;  fi;
    Syntax error: warning, this statement has no effect
    if i <> 0  then  k = 16/i;  fi;
                             ^ |

As you can see from  the  example this is useful for those  users who are
used to languages where '=' instead of '\:=' denotes assignment.

A sequence  of  one or more  statements is a statement sequence, and  may
occur  everywhere instead of  a single statement.  There  is nothing like
PASCAL\'s  BEGIN-END, instead each  construct is terminated by a keyword.
The most simple  statement sequence is a single  semicolon, which can  be
used as an empty statement sequence.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Assignments}%
\index{assignment!variable}

'<var> \:= <expr>;'

The *assignment* has the effect of assigning the value of the expressions
<expr> to the variable <var>.

The variable <var> may be an  ordinary variable (see "Variables"), a list
element selection '<list-var>[<int-expr>]' (see  "List Assignment")  or a
record  component   selection    '<record-var>.<ident>'   (see    "Record
Assignment").  Since a list element or a record component may itself be a
list or a record the  left hand side of  an assignment may be arbitrarily
complex.

Note that variables do  not have a type.  Thus  any value may be assigned
to any variable.   For example a  variable with  an integer  value may be
assigned a permutation or a list or anything else.

If the expression  <expr> is a  function  call then  this  function  must
return a value.   If the  function does not  return a  value  an error is
signalled and you enter a break loop  (see "Break Loops").   As usual you
can leave  the  break   loop   with  'quit;'.    If  you  enter   'return
<return-expr>;' the value of the expression  <return-expr> is assigned to
the variable, and execution continues after the assignment.

|    gap> S6 := rec( size := 720 );; S6;
    rec(
      size := 720 )
    gap> S6.generators := [ (1,2), (1,2,3,4,5) ];; S6;
    rec(
      size := 720,
      generators := [ (1,2), (1,2,3,4,5) ] )
    gap> S6.generators[2] := (1,2,3,4,5,6);; S6;
    rec(
      size := 720,
      generators := [ (1,2), (1,2,3,4,5,6) ] ) |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Procedure Calls}

'<procedure-var>();' \\
'<procedure-var>( <arg-expr> \{',' <arg-expr>\} );'

The   procedure  call   has   the   effect   of  calling   the  procedure
<procedure-var>.   A  procedure call is done exactly like a function call
(see "Function Calls").  The distinction between functions and procedures
is only for the  sake  of  the discussion,  {\GAP} does  not  distinguish
between them.

A *function* does return a value but does not produce a  side effect.  As
a convention the name of a function is a noun, denoting what the function
returns, e.g., 'Length', 'Concatenation' and 'Order'.

A *procedure* is a  function that does  not return  a value but  produces
some   effect.  Procedures  are called   only  for   this effect.  As   a
convention the name of a procedure is a verb, denoting what the procedure
does, e.g., 'Print', 'Append' and 'Sort'.

|    gap> Read( "myfile.g" );     & a call to the procedure Read
    gap> l := [ 1, 2 ];;
    gap> Append( l, [3,4,5] );    & a call to the procedure Append |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{If}%
\index{if}\index{if statement}%
\index{fi}\index{then}\index{else}\index{elif}

'if <bool-expr1>  then <statements1> \\
\{ elif <bool-expr2>  then <statements2> \} \\
{}[ else <statements3> ] \\
fi;'

The 'if' statement  allows  one  to  execute statements depending  on the
value of some boolean expression. The execution is done as follows.

First the expression <bool-expr1> following the 'if' is evaluated.  If it
evaluates to 'true' the statement sequence  <statements1> after the first
'then' is executed, and the execution of the 'if' statement is complete.

Otherwise the expressions <bool-expr2> following the 'elif' are evaluated
in turn.  There may  be any number of 'elif' parts, possibly none at all.
As soon  as an expression evaluates to 'true' the corresponding statement
sequence <statements2> is executed and execution of the 'if' statement is
complete.

If the 'if' expression  and all, if  any, 'elif' expressions  evaluate to
'false'  and there is  an 'else'  part, which is  optional, its statement
sequence <statements3>  is  executed   and the   execution of    the 'if'
statement is complete.  If there is no 'else' part  the 'if' statement is
complete without executing any statement sequence.

Since the 'if' statement is  terminated by the  'fi' keyword there  is no
question where an 'else' part belongs, i.e., {\GAP} has no dangling else.\\
In 'if <expr1>  then  if <expr2>  then <stats1>  else <stats2>  fi;  fi;'\\
the 'else' part  belongs  to  the  second   'if'  statement,  whereas  in\\
'if <expr1>  then  if  <expr2>  then  <stats1>  fi;  else  <stats2>  fi;'\\
the 'else' part belongs to the first 'if' statement.

Since an if statement is  not an expression it  is not possible to  write

|    abs := if x  > 0  then  x;  else  -x;  fi;|

which would,  even  if  legal  syntax, be   meaningless,  since  the 'if'
statement does not produce a value that could be assigned to 'abs'.

If one expression evaluates neither to 'true'  nor to 'false' an error is
signalled and a break loop (see "Break Loops") is  entered.  As usual you
can leave   the break loop  with 'quit;'.   If you  enter 'return true;',
execution  of the  'if' statement  continues  as if the expression  whose
evaluation    failed had evaluated to  'true'.     Likewise, if you enter
'return  false;', execution of  the  'if' statement  continues as if  the
expression whose evaluation failed had evaluated to 'false'.

|    gap> i := 10;;
    gap> if 0 < i  then
    >        s := 1;
    >    elif i < 0  then
    >        s := -1;
    >    else
    >        s := 0;
    >    fi;
    gap> s;
    1        # the sign of i |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{While}%
\index{while}\index{while loop}\index{loop!while}

'while <bool-expr>  do <statements>  od;'

The  'while' loop executes the statement  sequence <statements> while the
condition <bool-expr> evaluates to 'true'.

First <bool-expr> is evaluated.  If it evaluates  to 'false' execution of
the 'while' loop  terminates and the  statement immediately following the
'while' loop is executed next.  Otherwise if  it evaluates  to 'true' the
<statements> are executed and the whole process begins again.

The  difference  between the 'while'   loop and the  'repeat  until' loop
(see "Repeat") is that the <statements>  in  the  'repeat until' loop are
executed at least  once, while the  <statements> in the  'while' loop are
not executed at all if <bool-expr> is 'false' at the first iteration.

If  <bool-expr>  does  not  evaluate to 'true'  or 'false'  an  error  is
signalled  and a break loop (see "Break Loops") is entered.  As usual you
can  leave the break loop with 'quit;'.   If you  enter  'return false;',
execution  continues with the  next  statement immediately following  the
'while'  loop.   If you  enter  'return  true;',  execution continues  at
<statements>, after  which  the  next evaluation of <bool-expr> may cause
another error.

|    gap> i := 0;;  s := 0;;
    gap> while s <= 200  do
    >        i := i + 1;  s := s + i^2;
    >    od;
    gap> s;
    204        # first sum of the first i squares larger than 200 |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Repeat}%
\index{repeat}\index{repeat loop}\index{loop!repeat}\index{until}

'repeat <statements>  until <bool-expr>;'

The 'repeat' loop executes  the statement sequence <statements> until the
condition <bool-expr> evaluates to 'true'.

First <statements> are executed.   Then <bool-expr> is  evaluated.  If it
evaluates   to 'true' the  'repeat'   loop  terminates and  the statement
immediately following the  'repeat' loop is  executed next.  Otherwise if
it evaluates to 'false' the whole process begins again with the execution
of the <statements>.

The  difference  between the 'while'  loop (see "While")  and the 'repeat
until' loop  is that the  <statements>  in  the  'repeat until' loop  are
executed  at least once,  while the  <statements> in the 'while' loop are
not executed at all if <bool-expr> is 'false' at the first iteration.

If  <bool-expr>  does  not evaluate to  'true'   or  'false' a  error  is
signalled and a break loop (see "Break Loops") is entered.  As  usual you
can leave  the break loop  with 'quit;'.   If  you  enter 'return true;',
execution continues with the  next statement   immediately following  the
'repeat' loop.   If you  enter 'return   false;', execution continues  at
<statements>, after which the next  evaluation  of  <bool-expr> may cause
another error.

|    gap> i := 0;;  s := 0;;
    gap> repeat
    >        i := i + 1;  s := s + i^2;
    >    until s > 200;
    gap> s;
    204        # first sum of the first i squares larger than 200 |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{For}%
\index{for}\index{for loop}\index{loop!for}\index{do}\index{od}

'for <simple-var>  in <list-expr>  do <statements>  od;'

The 'for' loop executes   the statement sequence <statements>  for  every
element of the list <list-expr>.

The statement sequence  <statements> is  first executed with <simple-var>
bound  to  the first element of  the list  <list>, then with <simple-var>
bound to the second element of <list> and  so on.  <simple-var> must be a
simple  variable,   it   must not     be   a list     element   selection
'<list-var>[<int-expr>]'     or       a    record    component  selection
'<record-var>.<ident>'.

The execution of the 'for' loop is exactly equivalent to the 'while' loop

|    |<loop-list>| := |<list>|;
    |<loop-index>| := 1;
    while |<loop-index>| <= Length(|<loop-list>|) do
        |<variable>| := |<loop-list>|[|<loop-index>|];
        |<statements>|
        |<loop-index>| := |<loop-index>| + 1;
    od; |

with the   exception   that  <loop-list> and <loop-index>  are  different
variables for each 'for' loop that do not interfere with each other.

The list <list> is very often a range.\\
'for <variable> in [<from>..<to>] do <statements> od;'\\
corresponds to the more common\\
'for <variable> from <from> to <to>  do <statements> od;'\\
in other programming languages.

|    gap> s := 0;;
    gap> for i  in [1..100]  do
    >        s := s + i;
    > od;
    gap> s;
    5050 |

Note in the following example  how the modification of  the *list* in the
loop body causes the loop body also to be executed for the new values

|    gap> l := [ 1, 2, 3, 4, 5, 6 ];;
    gap> for i  in l  do
    >        Print( i, " " );
    >        if i mod 2 = 0  then Add( l, 3 * i / 2 );  fi;
    > od;  Print( "\n" );
    1 2 3 4 5 6 3 6 9 9
    gap> l;
    [ 1, 2, 3, 4, 5, 6, 3, 6, 9, 9 ] |

Note  in  the following example that  the  modification of the *variable*
that holds the list has no influence on the loop

|    gap> l := [ 1, 2, 3, 4, 5, 6 ];;
    gap> for i  in l  do
    >        Print( i, " " );
    >        l := [];
    > od;  Print( "\n" );
    1 2 3 4 5 6
    gap> l;
    [  ] |

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Functions}%
\index{function}\index{end}\index{local}%
\index{recursion}\index{recursive functions}%
\index{environment}\index{body}

|function (| [ <arg-ident> \{|,| <arg-ident>\} ] |)
    |[ |local   |<loc-ident> \{|, |<loc-ident>\} |;| ]|
    |<statements>|
end|

A function is in fact  a literal  and  not  a statement.  Such a function
literal can be  assigned to a variable  or to  a list element or a record
component.  Later  this function can  be called as described in "Function
Calls".

The following is an example  of a function definition.   It is a function
to compute values of the Fibonacci sequence (see "Fibonacci")

|    gap> fib := function ( n )
    >         local  f1,  f2,  f3,  i;
    >         f1 := 1;  f2 := 1;
    >         for i  in [3..n]  do
    >             f3 := f1 + f2;
    >             f1 := f2;
    >             f2 := f3;
    >         od;
    >         return f2;
    >     end;;
    gap> List( [1..10], fib );
    [ 1, 1, 2, 3, 5, 8, 13, 21, 34, 55 ] |

Because for  each of the formal arguments <arg-ident> and for each of the
formal locals <loc-ident>  a  new variable is allocated when the function
is called (see "Function  Calls"), it is  possible that a  function calls
itself.  This is usually called *recursion*. The following is a recursive
function that computes values of the Fibonacci sequence

|    gap> fib := function ( n )
    >         if n < 3  then
    >             return 1;
    >         else
    >             return fib(n-1) + fib(n-2);
    >         fi;
    >     end;;
    gap> List( [1..10], fib );
    [ 1, 1, 2, 3, 5, 8, 13, 21, 34, 55 ] |

Note that the recursive version needs '2 \*\ fib(<n>)-1' steps to compute
'fib(<n>)', while the  iterative version   of 'fib'  needs  only  '<n>-2'
steps.   Both are not optimal  however,  the library function 'Fibonacci'
only needs on the order of 'Log(<n>)' steps.

'<arg-ident> -> <expr>'

This  is a  shorthand for\\
'function (  <arg-ident> ) return <expr>; end'.\\
<arg-ident>  must  be  a single  identifier, i.e., it is  not possible to
write functions of several arguments this way.  Also 'arg' is not treated
specially,  so  it  is  also impossible to  write  functions that take  a
variable number of arguments this way.

The following is an example of a typical use of such a function

|    gap> Sum( List( [1..100], x -> x^2 ) );
    338350 |

When a function <fun1> definition  is evaluated  inside  another function
<fun2>, {\GAP}  binds all the identifiers inside the function <fun1> that
are identifiers of an argument or a  local of <fun2> to the corresponding
variable.  This set of bindings is called the environment of the function
<fun1>.  When <fun1> is called, its body is executed in this environment.
The  following implementation of a simple stack uses this.  Values can be
pushed  onto  the  stack  and  then  later  be  popped  off  again.   The
interesting thing here  is that the  functions 'push'  and  'pop'  in the
record returned by 'Stack' access the local variable 'stack' of  'Stack'.
When 'Stack' is called a new  variable  for  the  identifier  'stack'  is
created.   When the function  definitions of  'push' and  'pop' are  then
evaluated (as part of the  'return' statement)  each reference to 'stack'
is bound to this new variable.  Note also that the two stacks 'A' and 'B'
do not interfere, because each call of 'Stack' creates a new variable for
'stack'.

|    gap> Stack := function ()
    >         local   stack;
    >         stack := [];
    >         return rec(
    >             push := function ( value )
    >                 Add( stack, value );
    >             end,
    >             pop := function ()
    >                 local   value;
    >                 value := stack[Length(stack)];
    >                 Unbind( stack[Length(stack)] );
    >                 return value;
    >             end
    >         );
    >    end;;
    gap> A := Stack();;
    gap> B := Stack();;
    gap> A.push( 1 );  A.push( 2 );  A.push( 3 );
    gap> B.push( 4 );  B.push( 5 );  B.push( 6 );
    gap> A.pop();  A.pop();  A.pop();
    3
    2
    1
    gap> B.pop();  B.pop();  B.pop();
    6
    5
    4 |

This feature should be used rarely, since its implementation in {\GAP} is
not very efficient.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Return}%
\index{return}

'return;'

In this form 'return' terminates the call of  the innermost function that
is currently executing, and control returns to  the calling function.  An
error  is signalled if no  function is currently  executing.  No value is
returned by the function.

'return <expr>;'

In this form 'return' terminates the call of the innermost  function that
is currently executing, and returns  the value  of the expression <expr>.
Control returns  to the calling  function.  An error is  signalled if  no
function is currently executing.

Both statements  can also be used in  break loops  (see  "Break  Loops").
'return;'  has the effect  that the  computation  continues  where it was
interrupted by an error or the user hitting  '<ctr>C'.   'return <expr>;'
can be used to continue execution after an error.  What  happens with the
value <expr> depends on the particular error.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{The Syntax in BNF}%
\index{BNF}

This section contains the definition  of the {\GAP} syntax in Backus-Naur
form.

A  BNF is a set of rules, whose left side  is the name of  a  syntactical
construct.  Those  names  are enclosed in angle  brackets and  written in
<italics>.  The right side of each rule contains a possible form for that
syntactic  construct.   Each  right  side  may  contain  names  of  other
syntactic  constructs,  again  enclosed in angle brackets and written  in
<italics>,  or character  sequences that  must  occur literally; they are
written in 'typewriter style'.

Furthermore  each righthand side  can contain  the  following metasymbols
written in *boldface*.  If the right  hand  side contains forms separated
by a pipe symbol  ($\|$)  this means  that one  of the possible forms can
occur.  If a part of a form  is enclosed  in square brackets  ([ ])  this
means that this part is optional, i.e.  might be present or  missing.  If
part  of the form is enclosed  in curly braces  (\{  \})  this means that
the part may occur arbitrarily often, or possibly be missing.

\newpage
\begin{tabbing}
<Permutation>  \=\:= \= <Expr>                                      \kill
<Ident>        \>\:=  \>'a'$\|$...$\|$'z'$\|$'A'$\|$...$\|$'Z'$\|$'\_'
\{'a'$\|$...$\|$'z'$\|$'A'$\|$...$\|$'Z'$\|$'0'$\|$...$\|$'9'$\|$'\_'\}\\
<Var>          \>\:=  \><Ident>                                        \\
               \>$\|$ \><Var> '.' <Ident>                              \\
               \>$\|$ \><Var> '.' '(' <Expr> ')'                       \\
               \>$\|$ \><Var> '[' <Expr> ']'                           \\
               \>$\|$ \><Var> '\{' <Expr> '\}'                         \\
               \>$\|$ \><Var> '(' [ <Expr> \{ ',' <Expr> \} ] ')'      \\
<List>         \>\:=  \>'[' [ <Expr> ] \{',' [ <Expr> ] \} ']'         \\
               \>$\|$ \>'[' <Expr> [, <Expr> ] '..' <Expr> ']'         \\
<Record>       \>\:=  \>'rec(' [ <Ident> '\:=' <Expr>
                           \{',' <Ident> '\:=' <Expr> \} ] ')'         \\
<Permutation>  \>\:=  \>'(' <Expr> \{',' <Expr> \} ')'
                     \{ '(' <Expr> \{',' <Expr> \} ')' \}              \\
<Function>     \>\:=  \>'function (' [ <Ident> \{',' <Ident> \} ] ')'  \\
               \>     \>    [ 'local'  <Ident> \{',' <Ident> \} ';' ]  \\
               \>     \>    <Statements>                               \\
               \>     \>'end'                                          \\
<Char>         \>\:=  \>|'| <any character> |'|                        \\
<String>       \>\:=  \>'\"' \{ <any character> \} '\"'                \\
<Int>          \>\:=  \>'0'$\|$'1'$\|$...$\|$'9'
                     \{ '0'$\|$'1'$\|$...$\|$'9' \}                    \\
<Atom>         \>\:=  \><Int>                                          \\
               \>$\|$ \><Var>                                          \\
               \>$\|$ \>'(' <Expr> ')'                                 \\
               \>$\|$ \><Permutation>                                  \\
               \>$\|$ \><Char>                                         \\
               \>$\|$ \><String>                                       \\
               \>$\|$ \><Function>                                     \\
               \>$\|$ \><List>                                         \\
               \>$\|$ \><Record>                                       \\
<Factor>       \>\:=  \>\{'+'$\|$'-'\} <Atom> [ '\^' \{'+'$\|$'-'\} <Atom> ]\\
<Term>         \>\:=  \><Factor> \{ '\*'$\|$'/'$\|$'mod' <Factor> \}   \\
<Arith>        \>\:=  \><Term> \{ '+'$\|$'-' <Term> \}                 \\
<Rel>          \>\:=  \>\{ 'not' \} <Arith>
       \{ |=|$\|$|<>|$\|$|<|$\|$|>|$\|$|<=|$\|$|>=|$\|$|in| <Arith> \} \\
<And>          \>\:=  \><Rel> \{ 'and' <Rel> \}                        \\
<Log>          \>\:=  \><And> \{ 'or' <And> \}                         \\
<Expr>         \>\:=  \><Log>                                          \\
               \>$\|$ \><Var> [ '->' <Log> ]                           \\
<Statement>    \>\:=  \><Expr>                                         \\
               \>$\|$ \><Var> '\:=' <Expr>                             \\
               \>$\|$ \>'if'   <Expr>  'then' <Statements>             \\
               \>     \>\{ 'elif' <Expr>  'then' \=<Statements> \}     \\
               \>      \>[ 'else'                \><Statements>  ] 'fi'\\
               \>$\|$ \>'for' <Var> 'in' <Expr> 'do' <Statements> 'od' \\
               \>$\|$ \>'while' <Expr>  'do' <Statements>  'od'        \\
               \>$\|$ \>'repeat' <Statements>  'until' <Expr>          \\
               \>$\|$ \>'return' [ <Expr> ]                            \\
               \>$\|$ \>'quit'                                         \\
<Statements>   \>\:=  \>\{ <Statement> ';' \}                          \\
               \>$\|$ \>';'
\end{tabbing}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%E  Emacs . . . . . . . . . . . . . . . . . . . . . local Emacs variables
%%
%%  Local Variables:
%%  mode:               outline
%%  outline-regexp:     "\\\\Chapter\\|\\\\Section"
%%  fill-column:        73
%%  eval:               (hide-body)
%%  End:
%%



